Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-03-31 paper/metamonads.tex
some corrections

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but
it does not scale well, it is a debugging nightmare, and usually
suffers most from Software Aging \cite{parnas_aging}.  As with
other tasks in programming, the answer is to turn to statically
typed languages, which provide strong guarantees about the soundness
of the resulting code.  MetaOCaml in particular stands out: it
guarantees that both the code generator and the generated
code are well-typed.  While Template programming in C++ and 
Template Haskell do offer some amount of static typing 
\cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and MetaML.}
offers stronger guarantees.  As well, since MetaOCaml is implemented
on top of the very solid Ocaml language, one accrues additional benefits.

@@

And it \emph{almost} works. It is type correct.  When we look at the
generated code for |top|, we notice the `(ref (0, arr_1.(0)))' is
inlined wherever pivot was expected. For example, the swapping part of
the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in
...|. That is certainly not what we want. We meant for pivot searching
and swapping parts of the code to communicate via a shared mutable
state, variable |pv|. In the generated code, each piece of code gets
its own copy of |pv|. Rather then inline (ref (0, arr_1.(0)))
everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
and inlined |pv|. Incidentally, even in a pure functional code we
often wish to generate such a |let| statement so we can compute a
complex expression only once and use it value in several other
places. Again, we would rather not leave the job of detecting such
common subexpressions to the compiler (because in the complex code the
compiler may fail to detect the common subexpressions). The problem
with generating such |let| expressions is that we wish our geenrator
to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
of the |let|, which we would fill in later. In languages which
generate code in the string form this is easy. We generate the string
|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
representting the body, wcich would give us the complete |let|
statement. In MetaOCaml, we can't generate incomplete expressions, for
a good reason.

@@

The essence of our approach is assembling the code of the program by combining
instances of various aspects together. In the case of Gaussian
Elimination, the aspects are pivot determination, swapping rows and
columns, determinant tracking, etc. A primitive code generator yields a
code expression 
A combinator takes (complete and well-typed) code expressions combines
them into a composite expression (which is too statically guaranteed
to be well-formed and well-typed).

For example, the following is probably the simplest code generators:
\begin{code}
let one = .<1>.
\end{code}

(in the floating-poiint instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code fragments (perhaps say one more sentence about the brackets).
The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes, which are
...

@@

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the above code: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
simplest_code is not the function on integers -- it's a code
expression.

@@

However, examinine the code for |param_code1 plus one|
shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
we might have saved the computation on the code generation
stage. However, the generated code still exhibits the
duplication (which makes the code larger and may take longer to
execute, if the optimizer fails to eliminate the duplication).
So, we need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| so we could use
that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
But that seems impossible because |letgen exp| needs to generate
the let expression .<let t = exp in body>. but |letgen| doesn't have
the body yet. The body needs that temporary identifier .<t>. that will
be generated by |letgen| itself. Certainly |letgen| cannot generate
only a part of a let-expression. All generated expressions in
MetaOCaml are well-formed and complete.

>>>>

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml (cite website?) \cite{CTHL03} in particular stands
out: not only it guarantees the absense of problems such as
accidental variable capture -- so-called hygiene cite? \oleg{should we
  also mention automatic well-formedness}, MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
  MetaML.}  offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

@@

And it \emph{almost} works. It is type correct.  When we look at the
generated code for |top|, we notice the |(ref (0, arr_1.(0)))| is
inlined wherever pivot was expected. For example, the swapping part of
the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in ...|. 
That is certainly not what we want. We meant for pivot searching
and swapping parts of the code to communicate via a shared mutable
state, variable |pv|. In the generated code, each piece of code gets
its own copy of |pv|. Rather then inline |(ref (0, arr_1.(0)))|
everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
and inlined |pv|. Incidentally, even in a pure functional code we
often wish to generate such a |let| statement so we can compute a
complex expression only once and use it value in several other
places. Again, we would rather not leave the job of detecting such
common subexpressions to the compiler (because in the complex code the
compiler may fail to detect the common subexpressions). The problem
with generating such |let| expressions is that we wish our geenrator
to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
of the |let|, which we would fill in later. In languages which
generate code in the string form this is easy. We generate the string
|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
representting the body, wcich would give us the complete |let|
statement. In MetaOCaml, we can't generate incomplete expressions, for
a good reason.

@@

The essence of our approach is assembling the code of the program by
combining instances of various aspects together. In the case of
Gaussian Elimination, the aspects are pivot determination, swapping of
rows and columns, determinant tracking, etc. A primitive code
generator yields a code expression.  A combinator takes (complete and
well-typed) code expressions and combines them into a composite expression
(which is too statically guaranteed to be well-formed and well-typed).

We will be using MetaOcaml, which, as the instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them

For example, the following is probably the simplest code generator:
\begin{code}
let one = .<1>.
\end{code}

(in the floating-point instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code expressions (perhaps say one more sentence about the brackets).
The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes, which are
...

@@

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the above code: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
|simplest_code| is not the function on integers -- it's a code
expression.

@@

However, examinining the code for |param_code1 plus one|
shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
we might have saved the computation on the code generation
stage. However, the generated code still exhibits the
duplication (which makes the code larger and may take longer to
execute, if the optimizer fails to eliminate the duplication).
So, we need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| so we could use
that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
But that seems impossible because |letgen exp| needs to generate
the let expression .<let t = exp in body>. but |letgen| doesn't have
the body yet. The body needs that temporary identifier .<t>. that will
be generated by |letgen| itself. Certainly |letgen| cannot generate
only a part of a let-expression. All generated expressions in
MetaOCaml are well-formed and complete.
